---
title: "Introduction to linear regression"
author: ""
output:
  pdf_document: default
  html_document:
    includes:
      in_header: header.html
    css: ./lab.css
    highlight: pygments
    theme: cerulean
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(eval = TRUE, message = FALSE, warning = FALSE)
```

The Human Freedom Index is a report that attempts to summarize the idea
of "freedom" through a bunch of different variables for many countries
around the globe. It serves as a rough objective measure for the
relationships between the different types of freedom - whether it's
political, religious, economical or personal freedom - and other social
and economic circumstances. The Human Freedom Index is an annually
co-published report by the Cato Institute, the Fraser Institute, and the
Liberales Institut at the Friedrich Naumann Foundation for Freedom.

In this lab, you'll be analyzing data from Human Freedom Index reports
from 2008-2016. Your aim will be to summarize a few of the relationships
within the data both graphically and numerically in order to find which
variables can help tell a story about freedom.

## Getting Started

### Load packages

In this lab, you will explore and visualize the data using the
**tidyverse** suite of packages. The data can be found in the companion
package for OpenIntro resources, **openintro**.

Let's load the packages.

```{r load-packages, message=FALSE}
library(tidyverse)
library(openintro)
data('hfi', package='openintro')
```

### The data

The data we're working with is in the openintro package and it's called
`hfi`, short for Human Freedom Index.

1.  What are the dimensions of the dataset?

```{r}
?hfi
```

**The data set has 1458 observations and 123 variables.**

2.  What type of plot would you use to display the relationship between
    the personal freedom score, `pf_score`, and one of the other
    numerical variables? Plot this relationship using the variable
    `pf_expression_control` as the predictor. Does the relationship look
    linear? If you knew a country's `pf_expression_control`, or its
    score out of 10, with 0 being the most, of political pressures and
    controls on media content, would you be comfortable using a linear
    model to predict the personal freedom score?

**To display the relationship between the personal freedom score
(pf_score) and another numerical variable like pf_expression_control, we
can use a scatterplot. It helps us to visualize the relationship between
two numerical variables and is suitable for determining if a linear
relationship exists.**

```{r}
# Scatterplot of pf_score vs pf_expression_control
ggplot(hfi, aes(x = pf_expression_control, y = pf_score)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(title = "Relationship between Personal Freedom Score and Expression Control",
       x = "Expression Control Score",
       y = "Personal Freedom Score") +
  theme_minimal()
```

**The scatterplot shows us a positive relationship between the
Expression Control Score and the Personal Freedom Score. The data points
follow a linear trend.**

**A linear model can be used for the prediction purpose, but it should
be interpreted with caution, beccause we should consider the variability
in the data. It would be beneficial to assess the residuals to check if
the linear model is an adequate fit or if a more complex model such as
another non-linear approach might be needed to better capture the
relationship.**

If the relationship looks linear, we can quantify the strength of the
relationship with the correlation coefficient.

```{r cor}
hfi %>%
  summarise(cor(pf_expression_control, pf_score, use = "complete.obs"))
```

Here, we set the `use` argument to "complete.obs" since there are some
observations of NA.

## Sum of squared residuals

::: {#boxedtext}
In this section, you will use an interactive function to investigate
what we mean by "sum of squared residuals". You will need to run this
function in your console, not in your markdown document. Running the
function also requires that the `hfi` dataset is loaded in your
environment.
:::

Think back to the way that we described the distribution of a single
variable. Recall that we discussed characteristics such as center,
spread, and shape. It's also useful to be able to describe the
relationship of two numerical variables, such as `pf_expression_control`
and `pf_score` above.

3.  Looking at your plot from the previous exercise, describe the
    relationship between these two variables. Make sure to discuss the
    form, direction, and strength of the relationship as well as any
    unusual observations.

**Form:** The relationship appears to be linear, as the data points tend
to follow a straight-line pattern. This indicates that a linear model
could be suitable for capturing the relationship between
pf_expression_control and pf_score.

**Direction:** The scatterplot shows a positive relationship as its
moving from left to right in the upward direction, it means that as
pf_expression_control increases, the pf_score also tends to increase.
This suggests that countries with higher pf_expression_control scores
tend to have higher personal freedom scores.

**Strength:** While we can see a clear linear trend, along with that we
can also see some variability in the data, which means the strength of
the linear relationship is moderate rather than very strong. Some data
points are spread away from the line, indicating that the model might
not perfectly predict pf_score based solely on pf_expression_control.
But we observed from the above code that the correlation between the two
variables is 79%.

**Unusual Observations:** There are a few data points that lie further
away from the main cluster, which could be considered outliers. These
observations may suggest that certain countries do not follow the
general trend as closely as others, indicating unique situations or
factors affecting their freedom scores.

**Overall we can say that, the relationship is linear and positive, with
a moderate strength, and the presence of some variability and potential
outliers should be taken into account when interpreting predictions.**

Just as you've used the mean and standard deviation to summarize a
single variable, you can summarize the relationship between these two
variables by finding the line that best follows their association. Use
the following interactive function to select the line that you think
does the best job of going through the cloud of points.

```{r plotss-expression-score, eval=FALSE}
# This will only work interactively (i.e. will not show in the knitted document)
hfi <- hfi %>% filter(complete.cases(pf_expression_control, pf_score))
DATA606::plot_ss(x = hfi$pf_expression_control, y = hfi$pf_score)
```

After running this command, you'll be prompted to click two points on
the plot to define a line. Once you've done that, the line you specified
will be shown in black and the residuals in blue. Note that there are 30
residuals, one for each of the 30 observations. Recall that the
residuals are the difference between the observed values and the values
predicted by the line:

$$
  e_i = y_i - \hat{y}_i
$$

The most common way to do linear regression is to select the line that
minimizes the sum of squared residuals. To visualize the squared
residuals, you can rerun the plot command and add the argument
`showSquares = TRUE`.

```{r plotss-expression-score-squares, eval=FALSE}
DATA606::plot_ss(x = hfi$pf_expression_control, y = hfi$pf_score, showSquares = TRUE)
```

Note that the output from the `plot_ss` function provides you with the
slope and intercept of your line as well as the sum of squares.

4.  Using `plot_ss`, choose a line that does a good job of minimizing
    the sum of squares. Run the function several times. What was the
    smallest sum of squares that you got? How does it compare to your
    neighbors?

**The smallest sum of squares i got was 1449.45, as i am doing this lab
at home i compared it with my other experiments and i saw that the sum
of the squares range from 1400's to 16,000's**

## The linear model

It is rather cumbersome to try to get the correct least squares line,
i.e. the line that minimizes the sum of squared residuals, through trial
and error. Instead, you can use the `lm` function in R to fit the linear
model (a.k.a. regression line).

```{r m1}
m1 <- lm(pf_score ~ pf_expression_control, data = hfi)
```

The first argument in the function `lm` is a formula that takes the form
`y ~ x`. Here it can be read that we want to make a linear model of
`pf_score` as a function of `pf_expression_control`. The second argument
specifies that R should look in the `hfi` data frame to find the two
variables.

The output of `lm` is an object that contains all of the information we
need about the linear model that was just fit. We can access this
information using the summary function.

```{r summary-m1}
summary(m1)
```

Let's consider this output piece by piece. First, the formula used to
describe the model is shown at the top. After the formula you find the
five-number summary of the residuals. The "Coefficients" table shown
next is key; its first column displays the linear model's y-intercept
and the coefficient of `pf_expression_control`. With this table, we can
write down the least squares regression line for the linear model:

$$
  \hat{y} = 4.61707 + 0.49143 \times pf\_expression\_control
$$

One last piece of information we will discuss from the summary output is
the Multiple R-squared, or more simply, $R^2$. The $R^2$ value
represents the proportion of variability in the response variable that
is explained by the explanatory variable. For this model, 63.42% of the
variability in runs is explained by at-bats.

5.  Fit a new model that uses `pf_expression_control` to predict
    `hf_score`, or the total human freedom score. Using the estimates
    from the R output, write the equation of the regression line. What
    does the slope tell us in the context of the relationship between
    human freedom and the amount of political pressure on media content?

```{r}
# linear model
m2 <- lm(hf_score ~ pf_expression_control, data = hfi)

# summary of the model
summary(m2)

```

**Based on the output of the linear model, the equation of the
regression line can be written as:**

$$
\hat{hf\_score} = 5.1537 + 0.3499 \times pf\_expression\_control
$$

### Interpretation of the Slope

The slope of the regression line is approximately 0.3499. This indicates
that for every one-unit increase in the `pf_expression_control` score
(representing more political pressure and controls on media content),
the total human freedom score (`hf_score`) is expected to increase by
0.3499 units, on average.

The positive slope suggests that higher scores of
`pf_expression_control` (which indicate less political pressure on media
content) are associated with higher human freedom scores. Thus, as
political pressure and control on media content decrease, the overall
human freedom score tends to increase, indicating a positive
relationship between freedom of expression and general human freedom.

The $R^2$ value is 0.5775, that means 57.75% of the variability in the
total human freedom score is explained by the political pressure on
media content (`pf_expression_control`). This suggests a moderate fit of
the linear model to the data.

## Prediction and prediction errors

Let's create a scatterplot with the least squares line for `m1` laid on
top.

```{r reg-with-line}
ggplot(data = hfi, aes(x = pf_expression_control, y = pf_score)) +
  geom_point() +
  stat_smooth(method = "lm", se = FALSE)
```

Here, we are literally adding a layer on top of our plot. `geom_smooth`
creates the line by fitting a linear model. It can also show us the
standard error `se` associated with our line, but we'll suppress that
for now.

This line can be used to predict $y$ at any value of $x$. When
predictions are made for values of $x$ that are beyond the range of the
observed data, it is referred to as *extrapolation* and is not usually
recommended. However, predictions made within the range of the data are
more reliable. They're also used to compute the residuals.

6.  If someone saw the least squares regression line and not the actual
    data, how would they predict a country's personal freedom school for
    one with a 6.7 rating for `pf_expression_control`? Is this an
    overestimate or an underestimate, and by how much? In other words,
    what is the residual for this prediction?

To predict the personal freedom score (`pf_score`) for a country with a
`pf_expression_control` rating of 6.7 using the least squares regression
line equation:

### we use the regression equation

The equation for the regression line from the previous output is:

$$
\hat{pf\_score} = 4.61707 + 0.49143 \times pf\_expression\_control
$$

$$
\hat{pf\_score} = 4.61707 + 0.49143 \times 6.7
$$

Calculating this:

$$
\hat{pf\_score} = 4.61707 + 3.29258 = 7.90965
$$

So, the predicted personal freedom score for a country with a
`pf_expression_control` rating of 6.7 is approximately 7.91.

### Determining if it's an overestimate or an underestimate

To determine if this prediction is an overestimate or an underestimate,
we need the actual personal freedom score for a country with a
`pf_expression_control` of 6.7 for this we made the below csalculations:

```{r closet_match}

# Filtering the dataset to find the closest `pf_expression_control` value to 6.7
closest_match <- hfi %>%
  filter(abs(pf_expression_control - 6.7) < 0.1) %>%
  select(pf_expression_control, pf_score)

closest_match

```

```{r average_of_values}
#  average pf_score for the closest matches
average_pf_score <- closest_match %>%
  summarise(mean_pf_score = mean(pf_score))


average_pf_score

```

### Calculating the residual

The residual is calculated as:

$$
\text{Residual} = \text{Actual } pf\_score - \hat{pf\_score}
$$

$$
\text{Residual} = 8.01 - 7.90 = 0.11
$$

Since the residual is positive (0.11), it indicates that the predicted
value is an underestimate of the actual personal freedom score by
approximately 0.11 units.

## Model diagnostics

To assess whether the linear model is reliable, we need to check for (1)
linearity, (2) nearly normal residuals, and (3) constant variability.

**Linearity**: You already checked if the relationship between
`pf_score` and \`pf_expression_control' is linear using a scatterplot.
We should also verify this condition with a plot of the residuals vs.
fitted (predicted) values.

```{r residuals}
ggplot(data = m1, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  xlab("Fitted values") +
  ylab("Residuals")
```

Notice here that `m1` can also serve as a data set because stored within
it are the fitted values ($\hat{y}$) and the residuals. Also note that
we're getting fancy with the code here. After creating the scatterplot
on the first layer (first line of code), we overlay a horizontal dashed
line at $y = 0$ (to help us check whether residuals are distributed
around 0), and we also reanme the axis labels to be more informative.

7.  Is there any apparent pattern in the residuals plot? What does this
    indicate about the linearity of the relationship between the two
    variables?

\*\*In the plot we can see that, the residuals are plotted against the
fitted values (predicted `pf_score`). Here's what we can observe about
the plot:

### Analysis:

1.  **No Apparent Pattern in Residuals:**
    -   The residuals seem to be scattered randomly around the
        horizontal line at zero, without any clear pattern (e.g.,
        curves, clusters, or systematic structures).
    -   This randomness indicates that the residuals do not display any
        particular structure, which suggests that the linearity
        assumption holds.
2.  **Linearity Condition:**
    -   Since there is no apparent pattern, it suggests that the
        relationship between `pf_expression_control` and `pf_score` is
        approximately linear.

### Conclusion:

The residual plot indicates that the linear model fits the data well.
This suggests that a linear model is appropriate for predicting
`pf_score` using `pf_expression_control`.\*\*

**Nearly normal residuals**: To check this condition, we can look at a
histogram

```{r hist-res}
ggplot(data = m1, aes(x = .resid)) +
  geom_histogram(binwidth = 25) +
  xlab("Residuals")
```

or a normal probability plot of the residuals.

```{r qq-res}
ggplot(data = m1, aes(sample = .resid)) +
  stat_qq()
```

Note that the syntax for making a normal probability plot is a bit
different than what you're used to seeing: we set `sample` equal to the
residuals instead of `x`, and we set a statistical method `qq`, which
stands for "quantile-quantile", another name commonly used for normal
probability plots.

8.  Based on the histogram and the normal probability plot, does the
    nearly normal residuals condition appear to be met?

Based on the histogram and the normal probability plot (Q-Q plot), we
can make the following analysis:

### Histogram Analysis

1.  The histogram shows a symmetric shape. However, the bars are densely
    packed, making it difficult to clearly assess if the distribution is
    perfectly normal.

### Q-Q Plot Analysis

1.  The Q-Q plot shows that most of the residuals lie close to the
    straight line, which indicates that they follow a normal
    distribution reasonably well.
2.  There are some deviations at the tails, particularly at the extreme
    ends, suggesting that there may be some mild departure from
    normality (potentially a few outliers).

### Conclusion

Overall, the nearly normal residuals condition appears to be reasonably
met, as the residuals exhibit a roughly symmetric shape in the histogram
and follow the line closely in the Q-Q plot. There may be some mild
deviations from perfect normality at the extremes, but these do not seem
severe enough to significantly violate the assumption.

**Constant variability**:

9.  Based on the residuals vs. fitted plot, does the constant
    variability condition appear to be met?

**Analysis of the Residuals vs. Fitted Plot:**

**Spread of Residuals:** In the residuals vs. fitted, the residuals
appear to be scattered evenly around the horizontal line at zero, with
no obvious patterns or funnel-shaped distributions.

**Consistent Variability:** The spread of the residuals does not appear
to systematically increase or decrease across different levels of fitted
values. This suggests that the variability is roughly constant.

**Conclusion** Based on the residuals vs. fitted plot, the constant
variability condition seems to be met. The residuals show a consistent
spread across the range of fitted values, indicating that the assumption
is satisfied for the linear model.

\*\*

------------------------------------------------------------------------

## More Practice

-   Choose another freedom variable and a variable you think would
    strongly correlate with it.. Produce a scatterplot of the two
    variables and fit a linear model. At a glance, does there seem to be
    a linear relationship?

Here I am choosing the variables `hf_score` (Human Freedom Score) and
`ef_score` (Economic Freedom Score), as these two are likely to be
correlated since economic freedom is often associated with overall human
freedom.

```{r}

# Scatterplot of Human Freedom Score vs. Economic Freedom Score with a linear regression line
ggplot(hfi, aes(x = ef_score, y = hf_score)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  labs(title = "Scatterplot of Human Freedom Score vs. Economic Freedom Score",
       x = "Economic Freedom Score",
       y = "Human Freedom Score") +
  theme_minimal()

```

```{r}
# Fitting a linear model for Human Freedom Score as a function of Economic Freedom Score
model <- lm(hf_score ~ ef_score, data = hfi)

# Summary of the model
summary(model)

```

Based on the scatterplot and the linear model output, we can analyze the
relationship between `hf_score` (Human Freedom Score) and `ef_score`
(Economic Freedom Score):

### Scatterplot Analysis:

-   The scatterplot shows a strong positive relationship between
    `Economic Freedom Score` and `Human Freedom Score`. As the
    `ef_score` increases, the `hf_score` also tends to increase,
    indicating a linear relationship.
-   The data points are relatively close to the fitted line, suggesting
    a strong correlation.

### Linear Model Output:

-   **Regression Equation:** The equation for the regression line is: $$
    \hat{hf\_score} = 0.25906 + 0.99245 \times ef\_score
    $$ This means that for every one-unit increase in the
    `Economic Freedom Score`, the `Human Freedom Score` is expected to
    increase by approximately 0.99245 units, on average.

-   **Strength of the Relationship:** The $R^2$ value is 0.7308,
    indicating that about 73.08% of the variability in
    `Human Freedom Score` can be explained by the
    `Economic Freedom Score`. This is a relatively high $R^2$,
    suggesting a strong linear relationship between the two variables.

-   **Significance:** The p-value for the slope (`<2e-16`) is extremely
    small, indicating that the relationship between `ef_score` and
    `hf_score` is statistically significant.

### Conclusion:

There appears to be a strong linear relationship between the
`Economic Freedom Score` and the `Human Freedom Score`.

-   How does this relationship compare to the relationship between
    `pf_expression_control` and `pf_score`? Use the $R^2$ values from
    the two model summaries to compare. Does your independent variable
    seem to predict your dependent one better? Why or why not?

1.  **Model for `pf_expression_control` vs. `pf_score`:**
    -   The $R^2$ value was approximately **0.5775**, indicating that
        about **57.75%** of the variability in the
        `Personal Freedom Score` (`pf_score`) can be explained by the
        `Political Expression Control` (`pf_expression_control`).
2.  **Model for `ef_score` vs. `hf_score`:**
    -   The $R^2$ value was approximately **0.7308**, indicating that
        about **73.08%** of the variability in the `Human Freedom Score`
        (`hf_score`) can be explained by the `Economic Freedom Score`
        (`ef_score`).

### Comparison of the Relationships:

-   **Higher** $R^2$ for `ef_score` vs. `hf_score`: The $R^2$ value for
    the relationship between `Economic Freedom Score` and
    `Human Freedom Score` is higher (0.7308) compared to the $R^2$ value
    for the relationship between `Political Expression Control` and
    `Personal Freedom Score` (0.5775). This indicates that `ef_score` is
    a better predictor of `hf_score` than `pf_expression_control` is of
    `pf_score`.

-   **Explanation:**

    -   The higher $R^2$ value suggests that a larger proportion of the
        variation in `Human Freedom Score` can be explained by
        variations in the `Economic Freedom Score`. This implies a
        stronger linear relationship between `ef_score` and `hf_score`,
        meaning economic freedom may have a more direct impact on
        overall human freedom than political expression control does on
        personal freedom.

-   What's one freedom relationship you were most surprised about and
    why? Display the model diagnostics for the regression model
    analyzing this relationship.

**One freedom relationship that is most surprising for me is between
freedom of religion (pf_religion) and personal freedom (pf_score). While
it's known that countries with higher levels of religious freedom may
also enjoy higher personal freedom, the strength of this relationship
may not be as strong as anticipated due to the influence of other
factors on personal freedom.**

```{r}
# linear model: Freedom of religion vs. Personal freedom score
religion_model <- lm(pf_score ~ pf_religion, data = hfi)

# Summary of the model
model_summary <- summary(religion_model)
print(model_summary)

# Scatterplot with regression line
ggplot(hfi, aes(x = pf_religion, y = pf_score)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(title = "Scatterplot of Personal Freedom Score vs. Freedom of Religion",
       x = "Freedom of Religion",
       y = "Personal Freedom Score") +
  theme_minimal()

# Residuals vs. Fitted plot
ggplot(data = religion_model, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs. Fitted Values",
       x = "Fitted values",
       y = "Residuals") +
  theme_minimal()

# Q-Q plot for normality of residuals
ggplot(data = religion_model, aes(sample = .stdresid)) +
  stat_qq() +
  stat_qq_line() +
  labs(title = "Q-Q Plot of Residuals") +
  theme_minimal()

# Histogram of residuals
ggplot(data = religion_model, aes(x = .resid)) +
  geom_histogram(binwidth = 0.5, fill = "grey", color = "black") +
  labs(title = "Histogram of Residuals",
       x = "Residuals",
       y = "Count") +
  theme_minimal()

```

The $R^2$ value for the linear model relating `pf_score` (Personal
Freedom Score) to `pf_religion` (Religious Freedom) is 0.2472. This
indicates that approximately 24.72% of the variability in the personal
freedom score can be explained by the religious freedom score.

The lower $R^2$ value here suggests that religious freedom is a less
strong predictor of personal freedom compared to economic freedom’s
relationship with human freedom.

The results indicate that while there is a positive relationship between
religious freedom and personal freedom, other factors may also play a
significant role in determining the personal freedom score.

------------------------------------------------------------------------
